\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-03-23 honors_thesis/journal_paper/paper.tex
Finished making changes based on feedback.  Removed token section

%% PSO Stuff
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\pers}{\ensuremath{P}}
\providecommand{\neigh}{\ensuremath{N}}
\providecommand{\nURand}{\ensuremath{U^\neigh}}
\providecommand{\pURand}{\ensuremath{U^\pers}}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\fppos}{\ensuremath{\Vec{fx}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{b}^\neigh}}
\providecommand{\fnbest}{\ensuremath{\Vec{fb}^\neigh}}
\providecommand{\pbest}{\ensuremath{\Vec{b}^\pers}}
\providecommand{\fpbest}{\ensuremath{\Vec{fb}^\pers}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\ncoeff}{\ensuremath{\phi^\neigh}}
\providecommand{\pcoeff}{\ensuremath{\phi^\pers}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
\providecommand{\ofunc}{\ensuremath{f}}
\providecommand{\swarm}{\ensuremath{swarm}}

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{V}}
\providecommand{\specpos}{\ensuremath{X}}
\providecommand{\specval}{\ensuremath{FX}}
\providecommand{\ln}{\ensuremath{ln}}
\providecommand{\rn}{\ensuremath{rn}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casedef}{\ensuremath{(\pbest:\nbest)}}
\providecommand{\casexn}{\ensuremath{(x:\neigh*)}}
\providecommand{\casexx}{\ensuremath{(x:x)}}
\providecommand{\casexl}{\ensuremath{(x:\ln)}}
\providecommand{\casexr}{\ensuremath{(x:\rn)}}
\providecommand{\casepn}{\ensuremath{(\pers*:\neigh*)}}
\providecommand{\casepl}{\ensuremath{(\pers*:\ln)}}
\providecommand{\casepr}{\ensuremath{(\pers*:\rn)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  i
%Even in desktop machines, systems with 4 or 8 cores are
%commonplace.  
In order to efficiently utilize these resources for
computationally intensive problems, algorithms need to run in parallel.

@@

This previous parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.

The purpose of this paper is to consider pso parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take on the order of hours.
Our purpose is to explore the question ``since pso is often run with
a swarm of size of 100, what would one do if 500 or a few thousand processors were available?''

For many functions there comes a point of diminishing returns with respect to adding particles.
!!!! TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! !!!!!
For the well-know benchmark function Sphere and as shown in \fig{basic-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
speed of convergence. In this case, the swam goes below $1*10^5$ in x iteration with 20 particles, but 
achieves this same value in  y iterations with a swarms size of 120, the additional 100 particles.
However consider the change going from 2000 particles to 2100 particles. Here the additional 100 particles
reduce the number of iterations to achieve $1*10^5$ from x iterations to y.
The effect is easier to see if we plot the value obtained as a function of swarm size, as show in 
fig{sideways-sphere} increasing the swarm size from 20 to 120 has a significant effect on the 
Other benchmark function behave similarly. As shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
Certainly these benchmark functions do not prove that this will be the case for all functions,
but it does provide some evidence that for at least some functions,
additional particles does not seem to have much benefit after some point,
some time earlier, some time later. This issue has been commented on by others
in the form of recommended swarms sizes for various types of functions (cite standard swarm paper).

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{evals_sphere.eps}
  \caption{Function Sphere with a swarm with various swarm sizes.}
  \label{fig:evals-sphere}
\end{figure}

@@

The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
explicitly on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.  This could not be done in gradient methods, for instance, where the
step size for sampling depends on the gradient and therefor the position and \emph{value}
at the current position, but it
has been done in simulated annealing~\cite{witte-tpds91}.

% Use this later?
% Skip this??
% 
% Speculative evaluation is possible in PSO because the motion of particles
% depends only on the \emph{positions} of the particle and its best neighbor, not
% the \emph{value} of the objective function at those positions.  We do not have
% to wait for a particle's function evaluation to complete in order to know where
% it might move to next, so we can compute the set of all possible next positions
% and send them to be evaluated along with the particle's current position.  Once
% the current position of each particle has been evaluated, we can determine
% which of the speculatively evaluated positions matches the branch taken by the
% particle.  We have thus evaluated two iterations at once.  This can easily get
% unwieldy if the set of next possible positions for each particle is large, but
% a wise choice of topology limits that set to a reasonable size.
% 
% But the following does need some intro

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle $i$ are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{i,t+1} &=
		\constriction \left[ \pvel_t +
			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
		\right] \\
\label{eq:posupdate}
	\ppos_{i,t+1} &= \ppos_{i,t} + \pvel_{i,t+1}
\end{align}
where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of independent
and identically distributed random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we have made the particle identifier $i$
and the time step $t$ or $t+1$ explicit in the subscript and moved other qualifiers,
such as the personal $\pers$ and personal best $\pers*$ qualifiers to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.
The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself, in that its own position is considered when updating the
particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.
% TOD USE THIS LATER!
% In this paper we use these
% topologies, as well as a parallel adaptation of the Complete topology, called
% Random, that has been shown to simulate the behavior of Complete with far less
% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
% randomly picks two other particles to share information with at each iteration,
% along with itself.  Thus in both the Ring and the Random topologies, all
% particles have 3 neighbors.

@@

The PSO algorithm just described can be parallelized by having each additional
processor add a particle or a set of particles to the swarm.
But as we have seen in the introduction, for some
functions, a more desirable use of resources is to speed up the algorithm,
performing more iterations instead of adding more particles (for a fixed number of function evaluations),
previous techniques have not addressed this issue.
In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.
We do this by refactoring the standard PSO equations as described here.

%First consider formula for the \emph{next} iteration of PSO, which is identical
%to Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time
%subscripts are incremented.  As noted above and
For simplicity in this
explanation, we restrict our discussion here to the case of a Ring topology
with two neighbors which we will call the ``right neighbor'' and ``left
neighbor'' neighbors identified for each particle $i$ as $\rn(i)$ and $\ln(i)$
respectively.

Next note that the computation of $\pbest$ and $\nbest$ is not seen in the
update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so we will make
this mathematically explicit.  This update is usually implemented as a pair of
independent ``if``-statements, but that is a bit awkward for our purpose here
so we instead use a set of indicator variables, one each representing each
possible update to the \emph{pair} $\casedef$.  These indicator variables allow
us to easily identify the cases in which updates will be needed for $\nbest$,
$\pbest$ or both.

For example, we can write $\casexn$ to identify the case where $\pbest$ will be
updated with the better position $\ppos_{i,t}$, shown in the first positions,
but a no new neighborhood best was found so $\nbest$ remains in the second
position.

For any topology with 2 neighbors there are 7 possible update cases listed in
Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
particle is also part of its own neighborhood (can cause an update to
$\nbest$). The list also omits the case where $\nbest$ is updated by
$\ppos_{i,t}$ but $\pbest$ was not updated, as this is logically impossible.
$\caseset$ represents the set of all cases.

For minimization, we can define this example indicator variable representing the case
where for particle $i$ will need to update $\pbest$ with the better
position $x_{i,t}$ for the next iteration ($t+1$), but $\nbest$ will not updated by:
\begin{align}
\nonumber
	\indic_{i,t+1}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
\nonumber
	& 1 \ if \  \ofunc(\ppos_{i,t+1}) > \ofunc(\pbest_{i,t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\ln(i),t+1}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\rn(i),t+1}) \\
\label{eq:deficasexn}
	& 0 \ otherwise
\end{align}

For compactness in our equations we represent this without the parameters since
they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.  The
indicator variables for the other update cases shown in Table~\ref{tab:evals}
($\casepn$, $\casepl$, $\casepr$, etc.) can be defined in similar fashion but
obviously using the corresponding boolean expression.

For each cases identified in Table~\ref{tab:evals} and identified by a
corresponding indicator variable defined above, we also provide notation for the \emph{next} velocity
update equation; once again using $\casexn$ as an example:
\begin{align}
\nonumber
	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1}, \ppos_{\ln(i),t+1}, \ppos_{\rn(i),t+1},
        \pbest_{i,t} , \nbest_{i,t}, \pURand_{i,t}, \nURand_{i,t})= \\
\label{eq:defvocasexn}
		& \constriction \left[ \pvel_t +
			\pcoeff\pURand_{i,t+1}\otimes(\pbest_{i,t+1} - \ppos_{i,t}+1) +
			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
		\right]
\end{align}

Again we will drop the parameters for compactness and since
they can be inferred from the subscripts on $\specvel_{i,t}^{\casexn}$.
The update equation is taken exactly from the statdard update equations~\eqref{eq:velupdate}
and \eqref{eq:posupdate}, but
since it is restricted to a specific case, $\casexn$ for this example,
it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
$\pbest_{i,t+1}=\ppos_{i,t+1}$ and $\nbest_{i,t+1}=\nbest_{i,t+1}$ we can re-write
$\specvel_{i,t+1}^{\casexn}$ as:
\begin{align}
\nonumber
	\specvel_{i,t+1}^{\casexn} & = \\
\label{eq:defvcasexn}
		& \constriction \left[ \pvel_t +
			\pcoeff\pURand_{i,t+1}\otimes(\ppos_{i,t+1} - \ppos_{i,t+1}) +
			\ncoeff\nURand_{i,t+1}\otimes(\nbest_{i,t+1} - \ppos_{i,t+1})
		\right]
\end{align}

For each cases identified in
Table~\ref{tab:evals}
and identified by a corresponding indicator variable defined above,
we define also a position update equation with its compact form; once again using $\casexn$ as an example:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
\nonumber
	\specpos_{i,t+1}^{\casexn} & =  \ppos_{i,t} + \specvel_{i,t+1}^{\casexn}
\end{align}

Lastly, we also define notation for the value of the objective function at these points,
once again using $\casexn$ as an example:
\begin{align}
\label{eq:deffcasexn}
	\specval_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
\nonumber
	\specval_{i,t+1}^{\casexn} & =  \ofunc(\specval_{i,t+1}^{\casexn}
\end{align}

With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
\begin{align}
\label{eq:vel2update}
	\pvel_{i,t+1} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t+1}^{c} \\
\label{eq:pos2update}
	\ppos_{i,t+1} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
\end{align}

The value of the objective function at $\ppos_{i,t+1}$ is given by:
\begin{align}
\label{eq:val2update}
	\ofunc (\ppos_{i,t+1}) = \sum_{c \in \caseset} \ \specval_{i,t+1}^{c} \ \indic_{i,t+1}^{c}
\end{align}
or, with $\specpos$ and $\indic$ written with the parameters which were omitted above:
\begin{align}
\nonumber
\ofunc & (\ppos_{i,t+1}) = \sum_{c \in \caseset} \\
\nonumber
& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1},\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1},
\pbest_{i,t},\nbest_{i,t},\pURand_{i,t}, \nURand_{i,t})) \\
\label{eq:val2updatelong}
& \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
\end{align}

None of the $\specpos_{i,t+1}^{c} \ \ c \in\ \caseset$ depend upon $f(i,x_t)$
or any other objective function evaluation at $t+1$.
Thus PSO has been re-factored such that the algorithm can begin computing
all of the objective function evaluations needed in iteration $t+1$ before $f(x_t)$ is computed.

This refactoring comes a significant expense, instead of just computing
$\ofunc(\ppos_{i,t+1})$ and then $\ofunc(\ppos_{i,t+2})$ in the usual way, this refactored PSO
can compute $\ofunc(\ppos_{i,t+1})$ and all 
of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time.
Once all of the evaluations $f(\ppos_{i,t+1})$ are completed only one
of the indicator functions $\indic_{i,t}^{c} \ \ c \in \caseset$ will have been set to one;
the rest will be ignored, and might just as well never have been computed.
We call these evaluated points $\specpos_{i,t+1}^{c}$ ``speculative children''
because they may or may not be needed.

The potential value of these speculative children lies in the phase used above
``this re-factored PSO
can compute $\ofunc(\ppos_{i,t+1})$ and all 
of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
Remember that our intent is to accelerate parallel PSO and that we
are interested in objective functions that have at least non-trivial computational
complexity.
Thus we assume that computing
$\ofunc(\ppos_{i,t+1})$ and all 
of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ is the computationally difficult part.

Suppose that unlimited processors were available and that the evaluation of 
the objective function took 1 hour. Suppose also that for each particle $i$ 
the computations
$\ofunc(\ppos_{i,t+1})$ and all 
of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ were sent to separate
processors and completed in parallel.
\emph{Given} the values 
of $\ofunc(\ppos_{i,t+1})$ and all 
$\ofunc(\ppos_{i,t+1}) i \in \swarm$ 
the computation of the indicator functions $indic_{i,t}^{c} \ \ c \in \caseset$ and to complete iteration t+1
is inconsequential relative to the computational cost of obtaining the function values. 
In this environment all of the function evaluations needed for 2 iterations of PSO
could be completed in just one hour where as the conventional implementation would
require two hours for those same 2 iterations of PSO.

While the algebra show that the computation of 
equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} (at iteration $t+2$),
There are some issues better seen in an algorithmic setting, in particular the use of the random
vectors, 
$\pURand_{i,t}$ and $\nURand_{i,t}$. As with the algebra each transformation of the algorithm is
provably equivalent (except that the algorithms will no longer be able to do an odd number of iteration,
since it will do two at a time). At this point we are moving work around and adding work, but the
resultant algorithm, produces numerically identical results.

\begin{algorithm} [h!]
  \caption{Standard PSO (minimizing)}
  \label{alg:standard}
  \begin{algorithmic}[1]
    \STATE t=0
    \STATE Initialize the swarm including
    \FOR{$i \in \swarm$}
      \STATE {$\ppos_{i,t} \leftarrow Random position$}
      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
    \ENDFOR
    \REPEAT
        \STATE $t=t+1$
        \FOR{$i \in \swarm$}
          \STATE {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$}
        \ENDFOR
        \FOR{$i \in \swarm$}
          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
          \label{alg1:startifs}
          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
          \ENDIF
          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
          \ENDIF
          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
          \ENDIF
          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
          \label{alg1:endifs}
          \ENDIF
          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
        \ENDFOR
    \UNTIL {$t >$ max iteration or convergence criteria met}
  \end{algorithmic}
\end{algorithm}

This algorithm can be transformed, will retaining it exact behavior, by
``unrolling'' one iteration of the ``Repeat'' loop, re-writing the ``if'''s in lines
\ref{alg1:startifs} through
\ref{alg1:endifs} to use the indicator functions $\indic_{i,t+1}$, and

\begin{algorithm} [h!]
  \caption{Unrolled PSO}
  \label{alg:unrolled}
  \begin{algorithmic}[1]
    \STATE t=0
    \STATE Initialize the swarm including
    \FOR{$i \in \swarm$}
      \STATE {$\ppos_{i,t} \leftarrow Random position$}
      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
    \ENDFOR
    \REPEAT
        \STATE $t=t+1$
        \FOR{$i \in \swarm$}
          \label{alg2:fatitert}
          \STATE DISPATCH {$\fppos_{i.t} \leftarrow \ofunc(\ppos_{i,t})$} to a seperate processor
        \ENDFOR
        \FOR{$i \in \swarm$}
          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
          \FOR{$c \in \caseset$}
            \STATE Compute $\specvel_{i,t+1}^c$ and $\specpos_{i,t+1}^c$.
	    Examples shown in equations~\eqref{eq:defvcasexn} and~\eqref{eq:defpcasexn}
            \label{alg2:fatitertp1}
            \STATE DISPATCH computation of $\specval_{i,t+1}^c$ to a seperate processor.
	    Example shown in equation~\eqref{eq:deffcasexn}.
	    \COMMENT Note that none of the computations, $\specvel_{i,t+1}^c$,
	    $\specpos_{i,t+1}^c$ and $\specval_{i,t+1}^c$ depend on $\fppos_{i.t}$.
          \ENDFOR
        \ENDFOR
        \STATE WAIT for all function evaluations dispatched above to complete
        \FOR{$i \in \swarm$}
          \FOR{$c \in \caseset$}
            \STATE Compute $\indic_{i,t+1}^c$
	           Examples shown in equation~\eqref{eq:deficasexn} and~\eqref{eq:defpcasexn}
	  \ENDFOR 
          \STATE Compute $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:vel2update} 
          \STATE Compute $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:pos2update}
          \STATE Compute $\fppos_{i,t+1}$ as shown in equation~\eqref{eq:val2update}
        \ENDFOR

        \STATE $t=t+1$
        \FOR{$i \in \swarm$}
          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
          \STATE $\nURand_{i,t} \Leftarrow RandomVector()$
          \STATE $\pbest_{i,t}\leftarrow \pbest_{i,t-1}$
          \STATE $\fpbest_{i,t}\leftarrow \fpbest_{i,t-1}$
          \IF {$\fppos_{i,t}<\fpbest_{i,t-1}$}
            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
          \ENDIF
          \STATE $\nbest_{i,t}\leftarrow \nbest_{i,t-1}$
          \STATE $\fnbest_{i,t}\leftarrow \fnbest_{i,t-1}$
          \IF {$\fppos_{i,t}<\fnbest_{i,t}$}
            \STATE $\pbest_{i,t}\leftarrow \ppos_{i,t}$
            \STATE $\fpbest_{i,t}\leftarrow \fppos_{i,t}$
          \ENDIF
          \IF {$\fppos_{\ln(i),t}<\fnbest_{i,t}$}
            \STATE $\pbest_{i,t}\leftarrow \ppos_{\ln(i),t}$
            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\ln(i),t}$
          \ENDIF
          \IF {$\ofunc(\ppos_{i,t})<\ofunc(\nbest_{\rn(i),t})$}
            \STATE $\pbest_{i,t}\leftarrow \ppos_{\rn(i),t}$
            \STATE $\fpbest_{i,t}\leftarrow \fppos_{\rn(i),t}$
          \ENDIF
          \STATE Update $\pvel_{i,t+1}$ as shown in equation~\eqref{eq:velupdate} 
          \STATE Update $\ppos_{i,t+1}$ as shown in equation~\eqref{eq:posupdate}
        \ENDFOR
    \UNTIL {$t >$ max iteration or convergence criteria met}
  \end{algorithmic}
\end{algorithm}

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  Because a particle is also its own neighbor, one
of the $2(n+1)$ speculative positions can be eliminated, as a particle cannot
be the source of its $\nbest$ update while not updating its $\pbest$.  Thus we
have $2(n+1)-1$, or $2n+1$, evaluations.  Table~\ref{tab:evals} lists all
possible combinations of updates that can be seen by a particle with two
neighbors other than itself, which is the case in all of the topologies we use.
If all of these speculative positions are evaluated in parallel along with the
original, non-speculative position of the particle at time $t-1$, then one of
the speculative positions has to be the correct one, and two iterations have
been evaluated at once.

>>>>

%% PSO Stuff
\DeclareMathOperator{\URand}{U}
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\coeff}{\ensuremath{\phi}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}

%SpecExPSO Stuff
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  Even in desktop machines, systems with 4 or 8 cores are
commonplace.  In order to efficiently utilize these resources for
computationally intensive problems, algorithms need to run in parallel.

@@

All of these parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new subswarms.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.  However, there are many functions for which performing more
iterations with fewer particles is a better use of resources than performing
fewer iterations with more particles.

@@

The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
explicitly on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.  This could not be done in gradient methods, for instance, where the
step size for sampling depends on the gradient at the current position, but it
has been done in simulated annealing~\cite{witte-tpds91}.

Speculative evaluation is possible in PSO because the motion of particles
depends only on the \emph{positions} of the particle and its best neighbor, not
the \emph{value} of the objective function at those positions.  We do not have
to wait for a particle's function evaluation to complete in order to know where
it might move to next, so we can compute the set of all possible next positions
and send them to be evaluated along with the particle's current position.  Once
the current position of each particle has been evaluated, we can determine
which of the speculatively evaluated positions matches the branch taken by the
particle.  We have thus evaluated two iterations at once.  This can easily get
unwieldy if the set of next possible positions for each particle is large, but
a wise choice of topology limits that set to a reasonable size.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\nbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.
\( \coeff_1 \), \( \coeff_2 \), and \( \constriction \) are parameters with
prescribed values required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself, in that its own position is considered when updating the
particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.  In this paper we use these
topologies, as well as a parallel adaptation of the Complete topology, called
Random, that has been shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
randomly picks two other particles to share information with at each iteration,
along with itself.  Thus in both the Ring and the Random topologies, all
particles have 3 neighbors.

@@

The PSO algorithm just described can be parallelized by having each additional
processor add a particle or a set of particles to the swarm.  But for some
functions, a more desirable use of resources is to speed up the algorithm,
performing more iterations instead of adding more particles.  Barring
parallelizing the evaluation of the objective function itself, which may not be
possible, previous techniques have not addressed this issue.

In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.  If we can guess
what a particle's next position will be from its current position, we can
evaluate both in parallel and perform two iterations of PSO at once.  Lacking
a good way to guess a particle's next position, we can at least enumerate all
possible next positions of the particle, evaluate all of them at once, and one
of them will be right.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  Because a particle is also its own neighbor, one
of the $2(n+1)$ speculative positions can be eliminated, as a particle cannot
be the source of its $\nbest$ update while not updating its $\pbest$.  Thus we
have $2(n+1)-1$, or $2n+1$, evaluations.  Table~\ref{tab:evals} lists all
possible combinations of updates that can be seen by a particle with two
neighbors in addition to itself, which is the case in all of the topologies we
use.  If all of these speculative positions are evaluated in parallel along
with the original, non-speculative position of the particle at time $t-1$, then
one of the speculative positions has to be the correct one, and two iterations
have been evaluated at once.
